Measurement Clusterization in Bayesian D-optimal Designs in Infinite Dimensions

### 2024-07-21 latex/flat.tex
flat.tex updated but does not compile
flat.tex is the old paper. diff works and compiles (no pix tho)

% \documentclass[ba,preprint]{imsart}% use this for supplement article
\documentclass[ba]{imsart}
%\input{definitions}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}
%% \usepackage{amsaddr}

%% \theoremstyle{plain}
%% \newtheorem{theorem}{Theorem}
%% \newtheorem{corollary}[theorem]{Corollary}
%% \newtheorem{lemma}[theorem]{Lemma}
%% \newtheorem{proposition}[theorem]{Proposition}
%% \theoremstyle{definition}
%% \newtheorem{definition}[theorem]{Definition}
%% \newtheorem{example}[theorem]{Example}
%% \newtheorem{conjecture}[theorem]{Conjecture}
%% \theoremstyle{remark}
%% \newtheorem{remark}[theorem]{Remark}

\usepackage{thm-restate}
\usepackage{tikz,pgfplotstable}
\usetikzlibrary{patterns}
\pgfplotsset{compat=1.9} % set to 1.8 to get old behaviour
%\usepackage{hyperref}

\newcommand{\R}{\ensuremath{\mathbb{R}}}
\newcommand{\eps}{\varepsilon}
\newcommand{\M}{\mathcal{M}}

@@

% Definitions for second chapter
\newcommand{\data}{\mathbf{d}}
\newcommand{\param}{\mathbf{m}}
\newcommand{\normal}{\mathcal{N}}
\newcommand{\pr}{\mu_{\textup{pr}}} %Prior measure
\newcommand{\post}{\mu_{\textup{post}}^{\data, \obs}} % Posterior measure
\newcommand{\prmean}{\param_{\textup{pr}}} % Prior mean
\newcommand{\postmean}{\param_{\textup{post}}} % Posterior mean
\newcommand{\postcov}{\Gamma_{\textup{post}}} % Posterior covariance
\newcommand{\prcov}{\Gamma_{\textup{pr}}} % Prior covariance
\newcommand{\modcov}{\Gamma_{\textup{model}}} % Model covariance
\newcommand{\tmp}{\mathcal{G}}
\newcommand{\meas}{\mathbf{o}}
\newcommand{\ev}{\mathbf{e}} % eigenvector 
\newcommand{\func}{\mathbf{a}}
\newcommand{\tr}[1]{\textup{tr}\left \{#1 \right \} }
\newcommand{\ttr}[1]{\textup{tr}\ #1}
\newcommand{\rank}{\textup{rank}\ }
\newcommand{\des}{\eta} % vector of design parameters
\newcommand{\sigsqr}{\sigma^2}

%% %\usepackage{comment}% http://ctan.org/pkg/comment
%% %% %% %\excludecomment{proof}
%% %% \excludecomment{figure}
%% %% \let\endfigure\relax

%% %\overfullrule=0pt

%
\pubyear{2021}
\volume{TBA}
\issue{TBA}
% \doi{0000}
%\arxiv{}
\firstpage{1}
\lastpage{1}

@@

\startlocaldefs
% ** Local definitions **
\endlocaldefs

@@

%% *** Frontmatter ***

@@

%\title{\support{}}
\runtitle{Measurement Clusterization in D-optimal Designs}
%\author{Yair Daon}
%\numberwithin{equation}{section}

@@

\address[addr1]{Azrieli Faculty of Medicine, Bar-Ilan University, Safed, Israel
    \printead{e1} % print email address of "e1"
}

@@

\begin{abstract}
  %\input{abstract}
  Estimation of parameters in physical processes often demands costly
  measurements, prompting the pursuit of an optimal measurement
  strategy. Finding such strategy is termed the problem of
  \emph{optimal experimental design}, abbreviated as optimal
  design. Remarkably, optimal designs can yield tightly clustered
  measurement locations, leading researchers to fundamentally revise
  the design problem just to circumvent this issue. Some authors
  introduce error correlation among error terms that are initially
  independent, while others restrict measurement locations to a finite
  set of locations. While both approaches may prevent clusterization,
  they also fundamentally alter the optimal design problem.

  In this study, we consider Bayesian D-optimal designs, i.e.~designs
  that maximize the expected Kullback-Leibler divergence between
  posterior and prior. We propose an analytically tractable model for
  D-optimal designs over Hilbert spaces. In this framework, we make
  several key contributions: \textbf{(a)} We establish that
  measurement clusterization is a generic trait of D-optimal designs
  with independent Gaussian measurement errors, and prove that
  introducing correlations among measurement error terms mitigates
  clusterization. \textbf{(b)} We characterize D-optimal designs as
  reducing uncertainty across a subset of prior covariance
  eigenvectors. Finally, \textbf{(c)} We leverage this
  characterization to argue that measurement clusterization arises as
  a consequence of the pigeonhole principle: when more measurements
  are taken than there are locations where the select eigenvectors are
  large and others are small --- clusterization occurs.
  %
  %% In summary, in this study we shed light on an often ignored issue with
  %% Bayesian D-optimal designs. We characterize such designs and show why
  %% clusterization arisesit arisestheir inherent properties, propose
  %% strategies to address clusterization, and provide insights into the
  %% fundamental attributes underlying measurement design optimization.

\end{abstract}

%% ** Keywords **
\begin{keyword}[class=MSC]
\kwd[Primary ]{62F15}
\kwd{35R30}
\kwd[; secondary ]{28C20}
\end{keyword}

@@

%\input{intro}
\section{Introduction}\label{section:intro}
Measurements play a fundamental role in generating observations and
are indispensable to any inference process. Selecting the optimal set
of measurements holds particular significance when inferring
parameters of a physical process. Unlike many other fields, where
measurements are fixed, in this context, we have the freedom to choose
which measurements to take. This freedom should be taken advantage of:
measurements should be chosen to enhance accuracy, reduce costs, or
both. Whether measurements involve specifying electrode locations on
the skin in electric impedance tomography \cite{horesh2010impedance},
determining certain wavelengths in MRI \cite{horesh2008mri},
positioning sensors for detecting ground-reflected waves in the search
for oil \cite{horesh2008borehole}, or optimizing sensor locations in a
seismographic network \cite{rabinowitz1990, steinberg1995} --- the
selection of optimal measurements, referred to as the problem of
\emph{optimal design}, becomes crucial. Optimal measurements are
chosen according to specific \emph{design criteria}, with A- and
D-optimality being two of the most widely recognized and extensively
studied design criteria \cite{Chaloner1995,
  AlexanderianGloorGhattas14}.

@@

%% first: blue, second: red, third: green, fourth: black,
%% fifth:magenta
\begin{figure}
    \centering
    \includegraphics[height=0.5\textwidth]{example.pdf}
    \caption{Measurement clusterization for optimal designs when
      inverting for the initial condition of the 1D heat equation (see
      supplementary material for details). Measurement locations were
      chosen according to the Bayesian D-optimality criterion of
      Theorem \ref{thm:d optimality}. Measurement locations are
      plotted over the computational domain \(\Omega = [0, 1]\)
      (x-axis), for varying numbers of measurements (y-axis). The
      colored numbers are measurement indices, plotted for visual
      clarity. Measurement clusterization already occurs for three
      measurements: the second measurement (red) is overlaid with the
      third (green). For five measurements, first (blue) and second
      (red) measurements are clustered, as well as the fourth (black)
      and the fifth (magenta).}
  \label{fig:clusterization illustration}
\end{figure}

@@

Avoiding measurement clusterization is a pragmatic approach:
intuitively, researchers recognize that measurement clusterization is
undesirable, even though the underlying reasons may not be fully
clear. Consequently, they strive to prevent it and devise various
methodologies to avoid measurement clusterization. Yet each and every
one of these methodologies achieves its objective by imposing
restrictions on measurement locations, thereby fundamentally altering
the optimal design problem. To the best of my knowledge, no previous
study has tried to address some seemingly simple yet fundamental
questions:
%
Why does imposing correlations between observations alleviate
measurement clusterization?
%
Is measurement clusterization a generic phenomenon?
%
And, most importantly: Why does measurement clusterization occur?
%
%Should we aim to avoid measurement clusterization?
%
%Is it possible to substitute an optimal clustered design with an
%equally optimal non-clustered design?
%Can an optimal clustered design be relpaced with an equally optimal
%non-clustered design?

@@

\item \label{q:generic} \textbf{Is measurement clusterization a
  generic phenomenon?}
  %\subsection{An answer for Question \ref{q:generic}: Genericity of measurement clusterization}
  %% Computer implementation of Lemma \ref{thm:char} for the inverse
  %% problem outlined in Section \ref{section:how} also generates \(\obs\)
  %% as a solution to the D-optimal design problem.  Furthermore,
  Randomized numerical simulations of our model give rise to D-optimal
  designs that exhibit clusterization more than 95\% of the time (see
  code in supplementary material). Given our model's genericity, we
  expect measurement clusterization to be a generic phenomenon.

@@

  %% In Section \ref{section:vanishing}, we provide an insightful
  %% explanation for the optimality of clustered designs when no model
  %% error is present. We demonstrate that for our model, a D-optimal
  %% design measures only a small subset of prior eigenvectors, which are
  %% the prior eigenvectors with largest power spectrum. In real-life
  %% problems there are limited number of locations where: (a) the
  %% relevant prior eigenvectors are large, and other prior eigenvectors
  %% with large power in the prior spectrum are zero. Then measurement
  %% clusterization is a result of the pigeonhole principle, where there
  %% are more measurements than measurement locations satisfying (a) and
  %% (b).

  %% We conjecture that the prevalence of measurement clusterization
  %% arises due to the ease of discovering clustered designs.

%% \item \label{q:avoid} \textbf{Should we aim to avoid measurement
%%   clusterization?} Based on the analysis conducted in this study, we
%%   did not find any compelling reason to explicitly avoid optimal
%%   clustered designs.

%% \item \label{q:replace} \textbf{Is it possible to substitute an
%%   optimal clustered design with an equally optimal non-clustered
%%   design?} In Section \ref{section:vanishing}, we answer this question
%%   in the affirmative, although we show that numerical experiments
%%   conducted using our model indicate a strong preference for clustered
%%   designs.

@@

\begin{restatable}{theorem}{char}\label{thm:char}
  Let:
  \begin{itemize}
    \item The D-optimality design criterion
    \cite{AlexanderianGloorGhattas14}:
    \begin{align*}
      \begin{split}
        \tar(\obs) %:&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
        % 
        % 
        % 
        &= \frac12 \log \det ( I + \sigma^{-2} \prcov^{1/2} \fwd ^*
        \obs^* \obs \fwd \prcov^{1/2}), 
      \end{split}
    \end{align*}
  \item \(\obs\) a D-optimal design operator
    \begin{equation*}
      \obs = \argmax_{\|\meas_j\| = 1, j=1,\dots,m}\tar(\obs),
    \end{equation*}
  \item \(\{\lambda_i\}_{i=1}^\infty\) eigenvalues of
    \(\fwd\prcov\fwd^*\) in decreasing order of magnitude.
    %% \item \(\{\ev_i\}_{i=1}^\infty\) their corresponding eigenvectors.
  \item \(\{\eta_i\}_{i=1}^\infty\) eigenvalues of \(\obs^*\obs\).

@@

  Then:
  \begin{enumerate}
  \item  \(\tr{\obs^*\obs} = m\).
  \item \(\obs^*\obs\) and \(\fwd\prcov\fwd^*\) are simultaneously
    diagonalizable.
  \item \(k := \rank \obs^*\obs \leq m\) and
    \begin{equation*}
      \tar(\obs) = \frac12 \sum_{i=1}^{k} \log (1 + \sigma^{-2}\lambda_i\eta_i). %= \frac12 \sum_{i=1}^{m} \log (1 + \sigma^{-2}\lambda_i\eta_i).
    \end{equation*}
  %% \item 
  %%   \begin{equation*}
  %%     k = \argmax \left \{ k:\lambda_k^{-1} < \sigma^{-2}\frac{m}{k} + \frac{1}{k} \sum_{j=1}^{k}
  %%     \lambda_j^{-1} \right \}.
  %%   \end{equation*}
  \item
    \begin{equation*}
        \eta_i = \begin{cases}
          \frac{m}{k} - \sigma^2 \lambda_i^{-1} + \sigma^2 \frac{1}{k} \sum_{j=1}^k \lambda_j^{-1} & 1 \leq i \leq k \\
          0 & i > k 
        \end{cases}.
    \end{equation*}
  \item The covariance of the pushforwad \(\fwd_{*} \post\) is \(\left
    ( (\fwd \prcov \fwd^*)^{-1} + \sigma^{-2} \obs^*\obs \right
    )^{-1}\) and its eigenvalues are
    \begin{equation*}
      \theta_i =
      \begin{cases}
        \left(\frac{\sum_{j=1}^k \lambda_j^{-1} + \sigma^{-2}m}{k} \right )^{-1} & i \leq k \\
        \lambda_i &  i > k 
      \end{cases}
    \end{equation*}
  \end{enumerate}
%\end{theorem}
\end{restatable}

@@

%% Finally, in Lemma \ref{lemma:lax} we generalize a lemma for
%% calculating \(\frac{\der}{\der t} \log \det (I + X(t))\), where
%% \(X(t)\)is an operator valued function \cite{Lax07}.

\subsection{Limitations}\label{subsec:limitations}
The main limitation of this study is that our generic model does not
correspond to any specific real-life problem. It is generic enough to
be analytically tractable, but one may argue our model is too far
removed from any real application. To these claims I would answer that
scientists have a long history of studying models that are bare-bones
simplifications of real systems, e.g. the Ising model
\cite{cipra1987}, the Lorenz system \cite{brin}, the Lotka-Volterra
equations \cite{logan2006}, the Carnot engine \cite{kardar2007}, and
many others.

%\input{prelims}
\section{Preliminaries and Notation}\label{section:prelim}

@@

\begin{equation*}%\label{eq:O}
  \obs u = (\meas_1(u), \dots, \meas_m(u) )^t \in \R^m,\ u \in \hilo.
\end{equation*}

@@

\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps) (\obs \eps' +
      \eps)^t ]
    % 
    % 
    = \obs \modcov \obs^* + \sigma^2I_m , 
  \end{split}
\end{align}
where
\begin{align}\label{eq:modcov explained}
  \begin{split}
    [\obs \modcov \obs^*]_{ij} & = e_i^t \obs \modcov \obs^* e_j 
    %
    %
    %
    = \meas_i (\modcov \meas_j).% \text{ (by \eqref{eq:obs*})}.
  \end{split}
\end{align}
Taking \(\modcov = 0\) is a common practice
\cite{tarantola2005,Kaipio2005,Vogel02} and then \(\Sigma =
\sigma^2I_m\) is a scalar matrix which does not depend on \(\obs\).

@@

\begin{theorem}[Alexanderian, Gloor, Ghattas \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let \(\pr = \normal(\prmean,\prcov)\) be a Gaussian prior on \(\hilp\)
  and let \(\post = \normal(\postmean,\postcov)\) the posterior measure
  on \(\hilp\) for the Bayesian linear inverse problem \(\data = \obs
  \fwd\param + \obs \eps' + \eps\) discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}

@@

%% It is important to note that since \(\obs\) is finite-rank,
%% \(\prcov^{1/2} \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}\) is
%% trace-class.
\begin{definition}\label{def:d optimality}
  We say \(\obs^{\star}\) is \emph{D-optimal} if \(\obs^{\star} =
  \argmax_{\obs} \tar(\obs)\), where entries of \(\obs \in (\hilo^*)^m\)
  are constrained to some allowed set of measurements in \(\hilo^*\).
\end{definition}

Intuition for Theorem \ref{thm:d optimality} can be gained by
considering a Bayesian linear model in finite dimensions, with
Gaussian prior and Gaussian noise. Then, a D-optimal design minimizes
the determinant of the posterior covariance matrix, and this turns out
to be a regularized version of the frequentist D-optimality criterion
\cite{Chaloner1995}. Theorem \ref{thm:d optimality} and Definition
\ref{def:d optimality} carry a similar intuition:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \frac12 \log \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \text{ by \eqref{eq:objective}}\\
    &= \frac12 \log \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \frac12 \log \det \prcov \postcov^{-1} \text{ by \eqref{eq:postcov}}.
    %%  &= \frac12 \log \det \prcov -\frac12 \log \det \postcov.
  \end{split}
\end{align*}
We think of \(\prcov\) as constant, so a D-optimal design minimizes a
quantity analogous to the posterior covariance determinant, similarly
to the finite-dimensional case.

%% \subsection{Notation Summary}\label{subsec:notation}
%%   Our notation is summarized below. We let:
%%   \begin{itemize}
%%   \item \(\hilp, \hilo\) Hilbert spaces.
%%   \item \(\fwd:\hilp \to \hilo\) a linear compact operator.
%%   \item \(\pr \sim \mathcal{N}(0, \prcov)\) prior Gaussian measure on
%%     \(\hilp$, with prior covariance operator \(\prcov:\hilp \to \hilp$.
%%   \item \(\obs: \hilo \to \mathbb{R}^m\) measurement operator, where \(m
%%     \in \mathbb{N}\) is the number of measurements taken.
%%   \item \(\sigma^2 \in \mathbb{R}_{+}\) observation noise variance.
%%   \item \(\modcov\) model error covariance operator.
%%     %% \(\data = \obs \fwd \param + \eps$, where \(\eps \in
%%     %% \mathbb{R}^m\) isiid \(\mathcal{N}(0, \sigma^2)\) noise.
%%   \item \(\Sigma(\obs) = \obs \modcov \obs^* + \sigma^2I$. 
%%   \item \(\post\) the posterior measure, with covariance \(\postcov$.
%%   \item A D-optimality design criterion
%%     \cite{AlexanderianGloorGhattas14}:
%%     \begin{align*}
%%       \begin{split}
%%         \tar(\obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
%%         % 
%%         % 
%%         % 
%%         &= \frac12 \log \det ( I + \prcov^{1/2} \fwd ^* \obs^* \Sigma(\obs)^{-1} \obs
%%         \fwd \prcov^{1/2}).
%%       \end{split}
%%     \end{align*}
%%   %% \item \(\{\lambda_i\}_{i=1}^\infty\) eigenvalues of \(\fwd\prcov\fwd^*$
%%   %%   in decreasing order of magnitude.
%%   %% %% \item \(\{\ev_i\}_{i=1}^\infty\) their corresponding eigenvectors.
%%   %% \item \(\{\eta_i\}_{i=1}^\infty\) eigenvalues of \(\obs^*\obs$.
%%   \end{itemize}

%% \subsection{Sequential vs Simultaneous Optimization}\label{subsec:seq vs sim}
%% From defintion \ref{def:d optimality} we wish to characterize solution(s) of the
%% following optimization problem for \(\tar$. %%: (\hilo^*)^m \to \R$:
%% \begin{align}\label{eq:optimization}
%%   \obs^{\star} := \argmax_{\obs} \tar( \obs ) 
%%   = \argmax_{\obs} \frac12 \log \det 
%%   (I + \prcov^{1/2} \fwd^*\obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}),
%% \end{align}
%% where \(\obs\) is constrained to some allowed set of observations. We
%% call this problem ``simultaneous optimization'', since all
%% observations are decided on simulatneously.

%% For computational reasons, one may prefer to find the best
%% observations in a sequential manner. Denote
%% \begin{equation}\label{eq:def obs_k}
%%   \obs_k := (\meas_1,\dots, \meas_k)^t,  k\leq m.
%% \end{equation}
%% Sequential optimal design proceeds as follows. Find \(\meas_1\) by
%% maximizing \(\tar(\obs_1)$. Then, keeping \(\meas_1\) fixed --- find
%% \(\meas_2\) as the maximizer of \(\tar(\obs_2)$. Then, find \(\meas_3\) by
%% keeping \(\meas_1,\meas_2\) fixed and taking \(\meas_3\) as the maximizer
%% of \(\tar(\obs_3)$. Continue this way until \(\obs_m = \obs\) is
%% found, where \(m\) is the number of available observations. %% It is
%% %% important to notice that this scheme does not require actually
%% %% observing data --- in \eqref{eq:objective} data is averaged out.

%% The analysis in this paper is conducted for the general simultaneous
%% optimization case. The sequential optimization case is dealt with in
%% section \ref{subsec:clusterization sequential}. It is important to
%% note, however that all conclusions we arrive at for the simultaneous
%% case easily specialize to the sequential case by considering the
%% posterior as the next sequential step's prior.

%\input{constrained}
\section{The Constrained Optimization Problem of D-Optimal Design}\label{section:D and grad}
We seek a formulation of the D-optimal design problem via Lagrange
multipliers. We first find the gradient of $\tar$, then we suggest
unit-norm constraints on $\obs$ and find their gradients. Results of
this section are summarized in Theorem \ref{thm:constrained}. First,
recall that:
\begin{definition}\label{def:var}
  Let $F$ a real valued function of $\obs$. The first variation of $F$
  at $\obs$ in the direction $V$ is:
  \begin{equation*}
    \delta F(\obs) V := \frac{\der}{\der \tau}\Big |_{\tau=0}  F( \obs + \tau V).
  \end{equation*}

@@

%% Gradients are best thought of as row vectors. This will prove
%% important in section \ref{subsec:necessary}.

%% \subsection{The gradient of $\tar$}\label{section:objective}

>>>>

\documentclass[ba]{imsart}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{conjecture}[theorem]{Conjecture}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\usepackage{thm-restate}
\usepackage{tikz,pgfplotstable}
\usetikzlibrary{patterns}
\pgfplotsset{compat=1.9} 
\newcommand{\R}{\ensuremath{\mathbb{R}}}
\newcommand{\eps}{\varepsilon}
\newcommand{\M}{\mathcal{M}}

@@

\newcommand{\data}{\mathbf{d}}
\newcommand{\param}{\mathbf{m}}
\newcommand{\normal}{\mathcal{N}}
\newcommand{\pr}{\mu_{\textup{pr}}} \newcommand{\post}{\mu_{\textup{post}}^{\data, \obs}} \newcommand{\prmean}{\param_{\textup{pr}}} \newcommand{\postmean}{\param_{\textup{post}}} \newcommand{\postcov}{\Gamma_{\textup{post}}} \newcommand{\prcov}{\Gamma_{\textup{pr}}} \newcommand{\modcov}{\Gamma_{\textup{model}}} \newcommand{\tmp}{\mathcal{G}}
\newcommand{\meas}{\mathbf{o}}
\newcommand{\ev}{\mathbf{e}} \newcommand{\func}{\mathbf{a}}
\newcommand{\tr}[1]{\textup{tr}\left \{#1 \right \} }
\newcommand{\ttr}[1]{\textup{tr}\ #1}
\newcommand{\rank}{\textup{rank}\ }
\newcommand{\des}{\eta} \newcommand{\sigsqr}{\sigma^2}

 \pubyear{2021}
\volume{TBA}
\issue{TBA}
\firstpage{1}
\lastpage{1}

@@

\startlocaldefs
\endlocaldefs

@@

@@

\runtitle{Measurement Clusterization in D-optimal Designs}

@@

\address[addr1]{Azrieli Faculty of Medicine, Bar-Ilan University, Safed, Israel
    \printead{e1} }

@@

\begin{abstract}
  Estimation of parameters in physical processes often demands costly measurements, prompting the pursuit of an optimal measurement strategy. Finding such strategy is termed the problem of \emph{optimal experimental design}, abbreviated as optimal design. Remarkably, optimal designs can yield tightly clustered measurement locations, leading researchers to fundamentally revise the design problem just to circumvent this issue. Some authors introduce error correlation among error terms that are initially independent, while others restrict measurement locations to a finite set of locations. While both approaches may prevent clusterization, they also fundamentally alter the optimal design problem.

In this study, we consider Bayesian D-optimal designs, i.e.~designs that maximize the expected Kullback-Leibler divergence between posterior and prior. We propose an analytically tractable model for D-optimal designs over Hilbert spaces. In this framework, we make several key contributions: \textbf{(a)} We establish that measurement clusterization is a generic trait of D-optimal designs with independent Gaussian measurement errors, and prove that introducing correlations among measurement error terms mitigates clusterization. \textbf{(b)} We characterize D-optimal designs as reducing uncertainty across a subset of prior covariance eigenvectors. Finally, \textbf{(c)} We leverage this characterization to argue that measurement clusterization arises as a consequence of the pigeonhole principle: when more measurements are taken than there are locations where the select eigenvectors are large and others are small --- clusterization occurs. 
 \end{abstract}

\begin{keyword}[class=MSC]
\kwd[Primary ]{62F15}
\kwd{35R30}
\kwd[; secondary ]{28C20}
\end{keyword}

@@

\section{Introduction}\label{section:intro}
Measurements play a fundamental role in generating observations and
are indispensable to any inference process. Selecting the optimal set
of measurements holds particular significance when inferring
parameters of a physical process. Unlike many other fields, where
measurements are fixed, in this context, we have the freedom to choose
which measurements to take. This freedom should be taken advantage of:
measurements should be chosen to enhance accuracy, reduce costs, or
both. Whether measurements involve specifying electrode locations on
the skin in electric impedance tomography \cite{horesh2010impedance},
determining certain wavelengths in MRI \cite{horesh2008mri},
positioning sensors for detecting ground-reflected waves in the search
for oil \cite{horesh2008borehole}, or optimizing sensor locations in a
seismographic network \cite{rabinowitz1990, steinberg1995} --- the
selection of optimal measurements, referred to as the problem of
\emph{optimal design}, becomes crucial. Optimal measurements are
chosen according to specific \emph{design criteria}, with A- and
D-optimality being two of the most widely recognized and extensively
studied design criteria \cite{Chaloner1995,
  AlexanderianGloorGhattas14}.

@@

\begin{figure}
    \centering
    \includegraphics[height=0.5\textwidth]{example.png}
    \caption{Measurement clusterization for optimal designs when
      inverting for the initial condition of the 1D heat equation (see
      supplementary material for details). Measurement locations were
      chosen according to the Bayesian D-optimality criterion of
      Theorem \ref{thm:d optimality}. Measurement locations are
      plotted over the computational domain \(\Omega = [0, 1]\)
      (x-axis), for varying numbers of measurements (y-axis). The
      colored numbers are measurement indices, plotted for visual
      clarity. Measurement clusterization already occurs for three
      measurements: the second measurement (red) is overlaid with the
      third (green). For five measurements, first (blue) and second
      (red) measurements are clustered, as well as the fourth (black)
      and the fifth (magenta).}
  \label{fig:clusterization illustration}
\end{figure}

@@

Avoiding measurement clusterization is a pragmatic approach:
intuitively, researchers recognize that measurement clusterization is
undesirable, even though the underlying reasons may not be fully
clear. Consequently, they strive to prevent it and devise various
methodologies to avoid measurement clusterization. Yet each and every
one of these methodologies achieves its objective by imposing
restrictions on measurement locations, thereby fundamentally altering
the optimal design problem. To the best of my knowledge, no previous
study has tried to address some seemingly simple yet fundamental
questions:
Why does imposing correlations between observations alleviate
measurement clusterization?
Is measurement clusterization a generic phenomenon?
And, most importantly: Why does measurement clusterization occur?

@@

\item \label{q:generic} \textbf{Is measurement clusterization a
  generic phenomenon?}
Randomized numerical simulations of our model give rise to D-optimal
  designs that exhibit clusterization more than 95\% of the time (see
  code in supplementary material). Given our model's genericity, we
  expect measurement clusterization to be a generic phenomenon.

@@

@@

\begin{theorem}\label{thm:char}
  Let:
  \begin{itemize}
    \item The D-optimality design criterion
    \cite{AlexanderianGloorGhattas14}:
    \begin{align*}
      \begin{split}
        \tar(\obs) &= \frac12 \log \det ( I + \sigma^{-2} \prcov^{1/2} \fwd ^*
        \obs^* \obs \fwd \prcov^{1/2}), 
      \end{split}
    \end{align*}
  \item \(\obs\) a D-optimal design operator
    \begin{equation*}
      \obs = \argmax_{\|\meas_j\| = 1, j=1,\dots,m}\tar(\obs),
    \end{equation*}
  \item \(\{\lambda_i\}_{i=1}^\infty\) eigenvalues of
    \(\fwd\prcov\fwd^*\) in decreasing order of magnitude.
\item \(\{\eta_i\}_{i=1}^\infty\) eigenvalues of \(\obs^*\obs\).

@@

  Then:
  \begin{enumerate}
  \item  \(\tr{\obs^*\obs} = m\).
  \item \(\obs^*\obs\) and \(\fwd\prcov\fwd^*\) are simultaneously
    diagonalizable.
  \item \(k := \rank \obs^*\obs \leq m\) and
    \begin{equation*}
      \tar(\obs) = \frac12 \sum_{i=1}^{k} \log (1 + \sigma^{-2}\lambda_i\eta_i). \end{equation*}
\item
    \begin{equation*}
        \eta_i = \begin{cases}
          \frac{m}{k} - \sigma^2 \lambda_i^{-1} + \sigma^2 \frac{1}{k} \sum_{j=1}^k \lambda_j^{-1} & 1 \leq i \leq k \\
          0 & i > k 
        \end{cases}.
    \end{equation*}
  \item The covariance of the pushforwad \(\fwd_{*} \post\) is \(\left
    ( (\fwd \prcov \fwd^*)^{-1} + \sigma^{-2} \obs^*\obs \right
    )^{-1}\) and its eigenvalues are
    \begin{equation*}
      \theta_i =
      \begin{cases}
        \left(\frac{\sum_{j=1}^k \lambda_j^{-1} + \sigma^{-2}m}{k} \right )^{-1} & i \leq k \\
        \lambda_i &  i > k 
      \end{cases}
    \end{equation*}
  \end{enumerate}
\end{theorem}

@@

\subsection{Limitations}\label{subsec:limitations}
The main limitation of this study is that our generic model does not
correspond to any specific real-life problem. It is generic enough to
be analytically tractable, but one may argue our model is too far
removed from any real application. To these claims I would answer that
scientists have a long history of studying models that are bare-bones
simplifications of real systems, e.g. the Ising model
\cite{cipra1987}, the Lorenz system \cite{brin}, the Lotka-Volterra
equations \cite{logan2006}, the Carnot engine \cite{kardar2007}, and
many others.
 \section{Preliminaries and Notation}\label{section:prelim}

@@

\begin{equation*}\obs u = (\meas_1(u), \dots, \meas_m(u) )^t \in \R^m,\ u \in \hilo.
\end{equation*}

@@

\begin{align}\label{eq:Sigma}
  \begin{split}
    \Sigma(\obs) :&= \mathbb{E}[ (\obs \eps' + \eps) (\obs \eps' +
      \eps)^t ]
= \obs \modcov \obs^* + \sigma^2I_m , 
  \end{split}
\end{align}
where
\begin{align}\label{eq:modcov explained}
  \begin{split}
    [\obs \modcov \obs^*]_{ij} & = e_i^t \obs \modcov \obs^* e_j 
= \meas_i (\modcov \meas_j).\end{split}
\end{align}
Taking \(\modcov = 0\) is a common practice
\cite{tarantola2005,Kaipio2005,Vogel02} and then \(\Sigma =
\sigma^2I_m\) is a scalar matrix which does not depend on \(\obs\).

@@

\begin{theorem}[Alexanderian, Gloor, Ghattas \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let \(\pr = \normal(\prmean,\prcov)\) be a Gaussian prior on \(\hilp\)
  and let \(\post = \normal(\postmean,\postcov)\) the posterior measure
  on \(\hilp\) for the Bayesian linear inverse problem \(\data = \obs
  \fwd\param + \obs \eps' + \eps\) discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
&= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}

@@

\begin{definition}\label{def:d optimality}
  We say \(\obs^{\star}\) is \emph{D-optimal} if \(\obs^{\star} =
  \argmax_{\obs} \tar(\obs)\), where entries of \(\obs \in (\hilo^*)^m\)
  are constrained to some allowed set of measurements in \(\hilo^*\).
\end{definition}

Intuition for Theorem \ref{thm:d optimality} can be gained by
considering a Bayesian linear model in finite dimensions, with
Gaussian prior and Gaussian noise. Then, a D-optimal design minimizes
the determinant of the posterior covariance matrix, and this turns out
to be a regularized version of the frequentist D-optimality criterion
\cite{Chaloner1995}. Theorem \ref{thm:d optimality} and Definition
\ref{def:d optimality} carry a similar intuition:
\begin{align*}
  \begin{split}
    \tar(\obs) &= \frac12 \log \det ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}) \text{ by \eqref{eq:objective}}\\
    &= \frac12 \log \det \Big( \prcov ( \prcov^{-1} + \fwd ^* \obs^* \Sigma^{-1} \obs \fwd) \Big )\\
    &= \frac12 \log \det \prcov \postcov^{-1} \text{ by \eqref{eq:postcov}}.
\end{split}
\end{align*}
We think of \(\prcov\) as constant, so a D-optimal design minimizes a
quantity analogous to the posterior covariance determinant, similarly
to the finite-dimensional case.

 \section{The Constrained Optimization Problem of D-Optimal Design}\label{section:D and grad}
We seek a formulation of the D-optimal design problem via Lagrange
multipliers. We first find the gradient of $\tar$, then we suggest
unit-norm constraints on $\obs$ and find their gradients. Results of
this section are summarized in Theorem \ref{thm:constrained}. First,
recall that:
\begin{definition}\label{def:var}
  Let $F$ a real valued function of $\obs$. The first variation of $F$
  at $\obs$ in the direction $V$ is:
  \begin{equation*}
    \delta F(\obs) V := \frac{\der}{\der \tau}\Big |_{\tau=0}  F( \obs + \tau V).
  \end{equation*}

@@

\begin{proposition}\label{prop:tar grad}
  The gradient of the D-optimality objective $\tar$ is
  \begin{equation*}
\nabla \tar(\obs) = ( I - \modcov \obs^* \Sigma^{-1}\obs ) \fwd
    \postcov \fwd^* \obs^* \Sigma^{-1}
  \end{equation*}
\end{proposition}
